---
title: "Module 8"
author: "Ben Riccio"
date: "2023-12-01"
output: html_document
bibliography: Module8.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rgbif)
library(lme4)
library(MuMIn)
library(rnoaa)
library(data.table)
library(usmap)
library(car)
library(knitr)
library(kableExtra)
library(sf)
```
## Introduction 
This research focuses on the impact of climate change on the migration patterns of trans-Gulf migrants (TGMs), neotropical passerines traveling from Central and South America to breed in temperate North America each spring. The study addresses the critical trade-offs faced by these birds in choosing their arrival time, balancing the advantages of early mating opportunities with later arrival for increased food availability in seasonal ecosystems. Failure to adjust arrival dates in response to warmer temperatures may lead to population declines. 

Previous studies has suggested that te highest intensity of seasonal migration activity of TGMs occurred between late April and early May and average peak intensity occurred latest in the season at the westernmost places(@abbott2023inbound). To verify the previous findings on the TGMs migrated to MA, this project aims to leverage eBird, a popular citizen science platform, and meteorological data to analyze TGM arrival times in Massachusetts and investigate in the effect of of local weather condition on the TGM arrival times, which could potentially contributes to the broader understanding of avian biology in the context of climate change.

To be more specific, this project will focus specifically on these three questions:

1. Does arrival time vary according to temperature and wind variables along migration route for TGMs migrating to MA?

2. If arrival time does vary with meteorological conditions, what role will climate change potentially play in the population status of TGMs arriving in MA during the spring?

3. How does your analysis contribute to, challenge, or refine previous hypothesis concerning the role that climatic variables play in long-distance migration in passerine birds?

## Methods 

### Data acquiring
We first download eBird’s species occurrence data from the Global Biodiversity Information Facility (GBIF) and select 5 TGMs to analysis. Then we use rnoa to interact the data from NOAA’s National Climatic Data Center’s API to compile weather.

### Data Analysis
We first find the date when 25% of each of the TGM population have arrived in a given year as the arrival time for each species. Then we modeled this process as a logistics curve.

Aftert that we first identified the Julian day each year that corresponds to 25% of the population's arrival, and then analyzed how this arrival day varied across different years. In the preparation of our weather data, we adjusted the wind direction to a 180-degree scale and choose the columns of weather, which we want to work with. By combining two tables, each containing eBird and weather data, we could apply  linear mixed-effect modeling to assess the impact of weather parameters on arrival time. We first apply to  Anova() function to determine the significance of the fixed effects utilized in the model, and then we use the Dredge() function helped identify the optimal nested model that best fits our data.



```{r, bird data,message=FALSE,echo=FALSE,warning=FALSE}
species <- c("Setophaga ruticilla","Seiurus aurocapilla","Dolichonyx oryzivorus",
             "Icterus galbula","Hirundo rustica")

dat <- readRDS("massbird.data.RDS")
```

```{r,weather data, message=FALSE,echo=FALSE,warning=FALSE}
options(noaakey = "wwIFQCHBLJKxAcyoyPedBfumbhAQkGXP")

sts <- c(
  "GHCND:USW00013894", #Mobile, 
  "GHCND:USW00013881", #Charlotte, 
  "GHCND:USW00014739" #Boston
)
sta.d <- bind_rows( #bind the rows
  lapply(sts,function(x) ncdc_stations(stationid = x)$data ) 
  )%>%
  mutate(usmap_transform(.,input_names = c("longitude","latitude"),output_names = c("longitude.1", "latitude.1")))%>% 
  mutate(name=str_sub(name, -5,-4))%>%
  mutate(migr.day=c(10,5,0))%>% 
  separate(id,into = c("station.type","id"))


weather.d <- meteo_pull_monitors(sta.d$id,date_min = "2000-01-01")

```

```{r, bird data analysis,echo=FALSE,message=FALSE,warning=FALSE}
sp<- dat%>%
  group_by(year)%>%
  mutate(date=as.Date(paste0(year,"-",month,"-",day)),
         j.day=julian(date,origin=as.Date(paste0(unique(year),"-01-01")))
  )%>%
  group_by(species,year,j.day,date)%>%
  summarise(day.tot=sum(individualCount,na.rm=T))%>%
  group_by(species,year)%>%
  mutate(prop=cumsum(day.tot/sum(day.tot,na.rm = T)))%>%
  filter(year>1999)

sp.pred <- sp%>%
  group_by(year)%>%
 summarize(
   pred=predict(nls(prop~SSlogis(j.day,Asym, xmid, scal)),newdata=data.frame(j.day=min(j.day):max(j.day))),
   j.day=min(j.day):max(j.day),
  )%>%
  left_join(sp%>%dplyr::select(j.day,date))

sp.arrive.date <-sp.pred%>%
  group_by(year)%>%
  filter(j.day==j.day[which.min(abs(pred-0.25))])
sp.arrive.date

```

```{r, weather data analysis,message=FALSE,echo=FALSE,warning=FALSE} 
weather.d <- weather.d%>%
  mutate(
    year= as.integer(str_sub(date, 1 ,4)),
    date=as.Date(date)
    )%>%
  group_by(year)%>% 
 mutate(j.day=julian(date,origin=as.Date(paste0(unique(year),"-01-01"))),
  date2=date,
  wdir.rad=(180-abs(wdf2-180))*pi/180,
  wvec=cos(wdir.rad)*-1*awnd 
  )%>% 
  dplyr::select(id,year,date2,j.day,tmin,tmax,wvec)%>%
  left_join(sta.d%>%select(id,name,migr.day))%>% 
  mutate(j.day=j.day+migr.day)

weather.wk <-weather.d %>% 
  group_by(year,name) %>% 
  mutate(wk.tmin = frollmean(tmin, n=14,align="right"),
         wk.tmax = frollmean(tmax, n=14,align="right"),
         wk.wvec = frollmean(wvec, n=14,align="right")
         )%>%
  dplyr::select(j.day,date2,name,wk.tmin,wk.tmax,wk.wvec)

sp.arr.weath.d <- sp.arrive.date %>%
  left_join(weather.d)%>%
  left_join(sp%>%dplyr::select(year,date,j.day))

sp.arr.weath.wk <- sp.arrive.date %>%
  left_join(weather.wk)%>%
  left_join(sp%>%dplyr::select(year,date,j.day))


```


## Results 

```{r, modeling, message=FALSE ,echo=FALSE,warning=FALSE}

bird_model_d <- lmer(j.day~tmin*tmax*wvec+(1|name),sp.arr.weath.d,na.action = "na.fail")
bird_model_wk <- lmer(j.day~wk.tmin*wk.tmax*wk.wvec+(1|name),sp.arr.weath.wk,na.action = "na.fail")

arr_d_aic <- dredge(bird_model_d,fixed = c("tmin","tmax","wvec"),)
kb_d <- kable(arr_d_aic[1:4,],caption = "Fit values for nested models of the most complicated lme model for daily weather")
kable_styling(kb_d)
best_model_d <- lmer(j.day~tmin+tmax+wvec+(1|name),sp.arr.weath.d,na.action = "na.fail")
Anova(best_model_d)

arr_wk_aic <- dredge(bird_model_wk,fixed = c("wk.tmin","wk.tmax","wk.wvec"),)
kb_wk <- kable(arr_wk_aic[1:4,],caption = "Fit values for nested models of the most complicated lme model for two-week weather")
kable_styling(kb_wk)
best_model_wk <- lmer(j.day~wk.tmin+wk.tmax+wk.wvec+(1|name),sp.arr.weath.wk,na.action = "na.fail")
Anova(best_model_wk)

```
## Discussion 
Our results indicated that for both daily weather and two-week weather averages the best fitting model accounted for the effects of minimum temperature, maximum temperature, and wind vectors. The anova test results indicated that for the daily weather average, only min temperature may be significant. The two week average anova test showed wind vector and min temperature were significant.This indicates that weather does indeed have some effect on TMG sping arrival time to MA.

Climate change could therefore impact when birds arrive each spring. As weather conditions change due to climate change, this result predicts that birds will change their arrival time.

The @deppe2015fat paper found that going the other direction, winter arrival time in Mexico, did in fact depend on weather conditions, particularly humidity and wind conditions. While this paper examined TGM migration south, our data supports their findings. 

## Author Contributions 
Ben Riccio: Code, Results, Discussion. <br>
Huang Weihang: Intro, Methods.  

## References
